National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2023-09-05 analysis/paper/paper.Rmd
minor text changes and added UI paragraph to methods

---
title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou ranges in Canada, both survival and recruitment decrease with the percentage of the range that is disturbed. However, there is also variation in demographic rates among ranges, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships. Demographic projections derived from local range-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships. We present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections. We combine this model with methods for simulating local population dynamics and monitoring to show how monitoring requirements depend on landscape condition. Where anthropogenic disturbance is very high or very low, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At intermediate disturbance levels where uncertainty is higher, deploying few collars for more years may be a lower cost option for improving the reliability of status projections relative to more intensive short-term monitoring. We provide an R Shiny interface to enable others to explore the implications of alternative monitoring scenarios, disturbance trajectories, priors, and other model parameters for boreal caribou demographic projections. This tool can help users assess monitoring needs and strategies.

keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
---

@@

<!-- With the following code you can access and display values from the yml header above. -->
<!-- Link to data here: https://drive.google.com/file/d/179lw6MPK4l3xGmded9kofzJkB6VlJVb9/view?usp=sharing -->

@@

<!-- The actual document text starts here: -->
# Introduction

NOTE: find working version of intro and discussion [here:] (https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true).

<!--
Motivating example - forecasting impacts of disturbance - WBI and RoF example. In order to reduce uncertainty, need to integrate local information on current state of the population into forecasts.

@@

In Missisa (and other places, esp Ontario) available local data are old and limited. Not clear how indicative older data is of state of population. Need new data to resolve questions, but what type, how much is enough. Can't monitor in great detail everywhere all the time. More generally, need for tools to help assess the value of additional monitoring and the best way to invest additional monitoring resources, given the info that is already available. 
-->

# Methods

@@

Adult survival is estimated by tracking the fate of marked individuals (Kaplan-Meier ests) [brief description of field/logistical details - can get details from NBCKC Telemetry writeup (https://www.cclmportal.ca/sites/default/files/2021-10/Telemetry%20draft%20October%2027%202021.pdf), e.g. GPS allow for remote detection vs VHF-only require physical relocation of individual; collar sends out a
mortality signal when no longer active after a predetermined period of time; males only occasionally collared; age of collared individuals is usually unknown]

@@

For this investigation we assume collars are deployed in January, and last for up to 6 years. The target number of collared females $n$ varies among scenarios, as does the total duration of the monitoring program (Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be replaced each year ($o=1$), or there may be a 3 year gap between deployments ($o=4$). We focus on exploring a set of scenarios in which anthropogenic disturbance, measured as the percentage of area within 500 m of disturbance [@johnson_science_2020;@eccc_scientific_2011], increases over time from a variety of starting points (Table \@ref(tab:monitoringScenarios), Figure \@ref(fig:figDisturbance)). Scenarios in which anthropogenic disturbance does not increase over time are included in Supplement D.

<!--
*Monitoring WG input from BC, ON, NL, (MB pending): lots of detail provided but generally those numbers suggest we could try values of ~3-10 [e.g. 3,6,9?] for the cow multiplier, 5-50 for collars [e.g. 5,10,20,40,60?], [not sure about renewal interval, but see SaraM email for some insights]. We know (e.g. from the MonitoringWG perspectives report/update to Rettie summary) that number of years of monitoring can be as low as 2 or as many as 20+

@@

TO DO: run low initial population size scenario, put in appendix of one by one sensitivity checks? 
-->

## Simulation of local population dynamics and monitoring

@@

To simulate plausible plausible demographic trajectories for example populations we used a modified version of Johnson et al's [-@johnson_science_2020] two-stage demographic model described by Dyson et al. [-@dyson_existing_2022]. The true number of post-juvenile females that survive from year $t$ to the census $\dot{W}_t$ is binomially distributed with true survival probability $\dot{S}_t$: $\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum potential recruitment rate is adjusted for sex ratio, misidentification biases (Supplement B) and (optionally) delayed age at first reproduction [@eacker_web-based_2019;@decesare_estimating_2012]: $$\dot{X}_t=\frac{c\dot{R}_t/2}{1+c\dot{R}_t/2};c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}, $$
where $w$ is a multiplier that defines the apparent number of adult females in the composition survey as a function of the number of collared animals, $q$ is the ratio of young bulls to adult females, $u$ is the probability of misidentifying young bulls as adult females and vice versa, and $z$ is the probability of missing a calf.

@@

The number of cows in the composition survey $\hat{W}_t$ is given by the number of collared cows $\hat{T}_t$ and the apparent number of adult females per collared female calves observed in the composition survey $w$: 
$$\hat{W}_t = w\hat{T}_t.$$
The number of observed calves $\hat{J}_t$ also depends on the unadjusted apparent recruitment for the population $\dot{R}_t$:
$$\hat{J}_t \sim \text{Binomial}(\hat{W}_t,\dot{R}_t).$$

@@

We model recruitment probability as a function of anthropogenic disturbance and fire with a log link [@johnson_science_2020;@stewart_climate-informed_2023;@dyson_existing_2022] and Gaussian distributed random year effect [@eacker_web-based_2019]: $$\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).$$ 
Estimated recruitment probability is adjusted for sex ratio, misidentification biases (Supplement B), and age of first reproduction [@eacker_web-based_2019;@decesare_estimating_2012] to get expected recruits per female:
$$X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.$$
The composition survey bias correction term $c$ depends on the apparent number of adult females per collared animal $w$, the ratio of young bulls to adult females $q$, the probability of misidentifying young bulls as adult females and vice versa $u$, and the probability of missing a calf $z$ (Supplement B). To maintain consistency with Johnson et al. [-@johnson_science_2020] we also provide an option to only adjust for sex ratio and misidentification biases $X_t=cR_t/2$.

@@

Also following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier estimates of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ are estimated from known-fate radio collar data using the survival R package [@therneau_modeling_2000], with precisions for years with no observed mortality estimated using a separate Bayesian model. Observed estimated survival is a Gaussian distributed function of the true survival probability: $$\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).$$ 
As in the national demographic model, we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link [@johnson_science_2020;@stewart_climate-informed_2023;@dyson_existing_2022] and Gaussian distributed random year effect [@eacker_web-based_2019]: $$S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).$$ 
To assess sensitivity of results to the survival estimation method and handle cases in which there is little or no survival data, we also implemented an alternative parametric exponential survival model, wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed [@schaub_integrated_2021]: $$\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).$$

All regression coefficients are assumed to be Gaussian distributed, with priors derived from the expected values and 95% confidence intervals estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We calibrated the prior distributions so that the 95% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5% and 97.5% quantiles of 1000 simulated survival and recruitment trajectories (Supplement C, Figure \@ref(fig:examplePredictions)).  
```{r priors, echo=FALSE,message=FALSE}

@@

TO DO: 
* note distribution of random effect parameters are truncated to ensure R and S remain within reasonable bounds (not guaranteed given log link). Not yet documented here.
* note Eacker survival method gives biased results in years when <12 months of survival data is available.
* Note I am dropping survival information for animals that are not alive at the beginning of a year to deal with censoring issues. There is probably some more elegant way to deal with this, but it won't make any difference as long as the collar deployment month is set to 1.
* consider & comment on alternative choices made in this example https://poissonconsulting.shinyapps.io/bboushiny/. Follow Eacker example in all particulars, change some aspects, or do comparisons?
* Ensure notation consistency amongst manuscript, UI, documentaiton and code where appropriate.

@@

@@

We should likely also have a pgph about costs of monitoring - it'll be pretty short, cause basically we have no idea what it costs since nobody will tell us :) Through the NBCKC Monitoring Working Group work, we tried over and over to get some basic cost estimates from P/T reps and practitioners, and hit a wall - nobody wanted to tell us a thing. Don't know if we'd now find anyone is more open to sharing, but don't think we should get our hopes up. But at the very least we'll want to acknowledge that we haven't accounted for relative costs of monitoring... [e.g. costs outlined in the Telemetry guidance report from the MonitoringWG: "Expenses to consider in radio-collared monitoring will vary, based on: helicopter (or other aircraft) use, purchase of tracking collars, data fees associated with the collars (the cost of which varies by operating system, recording frequency and data transmission frequency), and live-capture equipment (3 or more sets of capture nets and associated buckets, a netting gun, and ankle hobbles)"]
-->perhaps the keen MonitoringWG contacts would be willing to provide even really rough cost estimates, and/or even jsut a relative sense of how different it would be to e.g. monitor a few pops for many yrs vs lots of pops for a short time...?

>>>>

---
title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou ranges in Canada, both survival and recruitment decrease with the percentage of the range that is disturbed. However, there is also variation in demographic rates among ranges, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships. Demographic projections derived from local range-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships. We present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections. We combine this model with methods for simulating local population dynamics and monitoring to show how monitoring requirements depend on landscape condition. Where anthropogenic disturbance is very high or very low, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At intermediate disturbance levels where uncertainty is higher, deploying few collars for more years may be a lower cost option for improving the reliability of status projections relative to more intensive short-term monitoring. We provide an R Shiny interface to enable others to explore the implications of alternative monitoring scenarios, disturbance trajectories, priors, and other model parameters for boreal caribou demographic projections. This tool can help users assess monitoring needs and strategies.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

<!-- With the following code you can access and display values from the yml header above. -->

<!-- Link to data here: https://drive.google.com/file/d/179lw6MPK4l3xGmded9kofzJkB6VlJVb9/view?usp=sharing -->

@@

<!-- The actual document text starts here: -->

# Introduction

NOTE: find working version of intro and discussion [here:]
(<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true>).

```{=html}
<!--
Motivating example - forecasting impacts of disturbance - WBI and RoF example. In order to reduce uncertainty, need to integrate local information on current state of the population into forecasts.

@@

In Missisa (and other places, esp Ontario) available local data are old and limited. Not clear how indicative older data is of state of population. Need new data to resolve questions, but what type, how much is enough. Can't monitor in great detail everywhere all the time. More generally, need for tools to help assess the value of additional monitoring and the best way to invest additional monitoring resources, given the info that is already available. 
-->
```
# Methods

@@

Adult survival is estimated by tracking the fate of marked individuals
(Kaplan-Meier ests) [brief description of field/logistical details - can
get details from NBCKC Telemetry writeup
(<https://www.cclmportal.ca/sites/default/files/2021-10/Telemetry%20draft%20October%2027%202021.pdf>),
e.g. GPS allow for remote detection vs VHF-only require physical
relocation of individual; collar sends out a mortality signal when no
longer active after a predetermined period of time; males only
occasionally collared; age of collared individuals is usually unknown]

@@

For this investigation we assume collars are deployed in January, and
last for up to 6 years. The target number of collared females $n$ varies
among scenarios, as does the total duration of the monitoring program
(Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be
replaced each year ($o=1$), or there may be a 3 year gap between
deployments ($o=4$). We focus on exploring a set of scenarios in which
anthropogenic disturbance, measured as the percentage of area within 500
m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases over time from a variety of starting points (Table
\@ref(tab:monitoringScenarios), Figure \@ref(fig:figDisturbance)).
Scenarios in which anthropogenic disturbance does not increase over time
are included in Supplement D.

```{=html}
<!--
*Monitoring WG input from BC, ON, NL, (MB pending): lots of detail provided but generally those numbers suggest we could try values of ~3-10 [e.g. 3,6,9?] for the cow multiplier, 5-50 for collars [e.g. 5,10,20,40,60?], [not sure about renewal interval, but see SaraM email for some insights]. We know (e.g. from the MonitoringWG perspectives report/update to Rettie summary) that number of years of monitoring can be as low as 2 or as many as 20+

@@

TO DO: run low initial population size scenario, put in appendix of one by one sensitivity checks? 
-->
```
## Simulation of local population dynamics and monitoring

@@

To simulate plausible demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
two-stage demographic model described by Dyson et al.
[-@dyson_existing_2022]. The true number of post-juvenile females that
survive from year $t$ to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum
potential recruitment rate is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) delayed age at first reproduction
[@eacker_web-based_2019; @decesare_estimating_2012]:
$$\dot{X}_t=\frac{c\dot{R}_t/2}{1+c\dot{R}_t/2};c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}, $$
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $q$ is the ratio of young bulls to adult females, $u$
is the probability of misidentifying young bulls as adult females and
vice versa, and $z$ is the probability of missing a calf.

@@

The number of cows in the composition survey $\hat{W}_t$ is given by the
number of collared cows $\hat{T}_t$ and the apparent number of adult
females per collared female observed in the composition survey $w$:
$$\hat{W}_t = w\hat{T}_t.$$ The number of observed calves $\hat{J}_t$
also depends on the unadjusted apparent recruitment for the population
$\dot{R}_t$: $$\hat{J}_t \sim \text{Binomial}(\hat{W}_t,\dot{R}_t).$$

@@

We model recruitment probability as a function of anthropogenic
disturbance and fire with a log link [@johnson_science_2020;
@stewart_climate-informed_2023; @dyson_existing_2022] and Gaussian
distributed random year effect [@eacker_web-based_2019]:
$$\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).$$
Estimated recruitment probability is adjusted for sex ratio,
misidentification biases (Supplement B), and age of first reproduction
[@eacker_web-based_2019; @decesare_estimating_2012] to get expected
recruits per female:
$$X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.$$ The
composition survey bias correction term $c$ depends on the apparent
number of adult females per collared animal $w$, the ratio of young
bulls to adult females $q$, the probability of misidentifying young
bulls as adult females and vice versa $u$, and the probability of
missing a calf $z$ (Supplement B). To maintain consistency with Johnson
et al. [-@johnson_science_2020] we also provide an option to only adjust
for sex ratio and misidentification biases $X_t=cR_t/2$.

@@

Also following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the survival R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
$$\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).$$ As in the national
demographic model, we model survival probability as an adjusted function
of buffered anthropogenic disturbance $A_t$ with a log link
[@johnson_science_2020; @stewart_climate-informed_2023;
@dyson_existing_2022] and Gaussian distributed random year effect
[@eacker_web-based_2019]:
$$S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).$$
To assess sensitivity of results to the survival estimation method and
handle cases in which there is little or no survival data, we also
implemented an alternative parametric exponential survival model,
wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed
[@schaub_integrated_2021]:
$$\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).$$

All regression coefficients are assumed to be Gaussian distributed, with
priors derived from the expected values and 95% confidence intervals
estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We
calibrated the prior distributions so that the 95% prior prediction
intervals for survival and recruitment from the Bayesian model match the
range between the 2.5% and 97.5% quantiles of 1000 simulated survival
and recruitment trajectories (Supplement C, Figure
\@ref(fig:examplePredictions)).

```{r priors, echo=FALSE,message=FALSE}

@@

TO DO: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
Ensure notation consistency amongst manuscript, UI, documentaiton and
code where appropriate.

@@

To make this work more user friendly and accessible to decision makers
we built a user interface that allows for the exploration of different
monitoring and anthropogenic disturbance scenarios. We have also made
the R code and functions needed to carry out this analysis available as
an R package caribouMetrics. We built the user interface as an R Shiny
app [@chang2023], building off and expanding an app published by Eacker
et al. [-@eacker_web-based_2019]. The app allows users to provide
existing local data or simulate data based on national models as shown
above, define an anthropogenic disturbance scenario and make
projections. The results shown are similar to Figure
\@ref(fig:examplePredictions) and allow for comparisons to the national
model. Many parameters controlling the Bayesian model, local data
simulation and antropogenic scenario are exposed in the app with
sensible defaults set, such that novice users can get reasonable results
quickly while advanced user can customise the projections to suite their
needs. The caribouMetrics package is available on GitHub at
<https://github.com/LandSciTech/caribouMetrics> and a package website
with detailed documentation is also available
(<https://landscitech.github.io/caribouMetrics/>).

@@

We should likely also have a pgph about costs of monitoring - it'll be
pretty short, cause basically we have no idea what it costs since nobody
will tell us :) Through the NBCKC Monitoring Working Group work, we
tried over and over to get some basic cost estimates from P/T reps and
practitioners, and hit a wall - nobody wanted to tell us a thing. Don't
know if we'd now find anyone is more open to sharing, but don't think we
should get our hopes up. But at the very least we'll want to acknowledge
that we haven't accounted for relative costs of monitoring... [e.g.
costs outlined in the Telemetry guidance report from the MonitoringWG:
"Expenses to consider in radio-collared monitoring will vary, based on:
helicopter (or other aircraft) use, purchase of tracking collars, data
fees associated with the collars (the cost of which varies by operating
system, recording frequency and data transmission frequency), and
live-capture equipment (3 or more sets of capture nets and associated
buckets, a netting gun, and ankle hobbles)"] --\>perhaps the keen
MonitoringWG contacts would be willing to provide even really rough cost
estimates, and/or even jsut a relative sense of how different it would
be to e.g. monitor a few pops for many yrs vs lots of pops for a short
time...?
